---
title: "07_review_assessment"
output: html_document
---

Load the packages you will need

```{r}

```


Read the "aquapark4_phytoplankton_2019.csv" data into R and assign it a variable name so you can work with it

```{r}

```

Print a summary of the data you read in above including the columns and their type (for example: numeric, character, date, etc)

```{r}

```

How many samples are diatoms or dinoflagellates? (Hint: we need to count each occurrence of them)

```{r}

```


Plot the cell abundance (as points) measured for every observation (y axis) over the entire year (x axis)

```{r}

```

Make a new table only containing observations of "Pseudo-nitzschia sp." counts and store it in a new variable

```{r}

```

How many times was it detected?

```{r}

```

Plot the measured "Pseudo-nitzschia sp." abundance over the year as points connected by lines. Add the alert threshold to the plot as a horizontal dashed, red line.

```{r}

```


Extra credit:

What DSP producing species were detected during 2019?

```{r}

```



- Add your name and today's date to the document heading

- Change the output type to word 

- Hide the messages that are printed when you load packages and read in the data file



