---
title: "07_review_assessment"
output: html_document
---

Load the packages you will need

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```


Read the "aquapark4_phytoplankton_2019.csv" data into R and assign it a variable name so you can work with it

```{r}
x <- read_csv("aquapark4_phytoplankton_2019.csv")
```

Print a summary of the data you read in above including the columns and their type (for example: numeric, character, date, etc)

```{r}
summary(x)
```
```{r}
glimpse(x)
```


How many samples are diatoms or dinoflagellates? (Hint: we need to count each occurrence of them)

```{r}
count(x, Groups)
```


Plot the cell abundance (as points) measured for every observation (y axis) over the entire year (x axis)

```{r}
ggplot(data=x, aes(x=date, y=`Cells/L`)) +
  geom_point()
```

Make a new table only containing observations of "Pseudo-nitzschia sp." counts and store it in a new variable

```{r}
pn <- filter(x, Specie == "Pseudo-nitzschia sp.")
```

How many times was it detected?

```{r}
pn
```

```{r}
count(pn, Specie)
```
```{r}
nrow(pn)
```


Plot the measured "Pseudo-nitzschia sp." abundance over the year as points connected by lines. Add the alert threshold to the plot as a horizontal dashed, red line.

```{r}
ggplot(data=pn, aes(x=date, y=`Cells/L`)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=40000, color="red", linetype="dashed")
```


Extra credit:

What DSP producing species were detected during 2019?

```{r}
dsp <- filter(x, `Toxin producing species` == "DSP")

group_by(dsp, Specie) |>
  summarise(observations = n())
```



- Add your name and today's date to the document heading 

- Change the output type to word 

- Hide the messages that are printed when you load packages and read in the data file 



