---
title: "THE NAMIBIAN SHELLFISH SANITATION PROGRAMME (NSSP) REVIEW REPORT FOR 2024/25"
output: word_document
---

```{r  message=FALSE, echo=FALSE}
source("../setup.R")
```


```{r message=FALSE, echo=FALSE}
x <- read_phyto_data()
```

```{r warning=FALSE}
y <- read_toxin_data()
```


```{r message=FALSE, echo=FALSE}
# make a new table format to use for the figures
z <- x |>
  select(-`SumOfTransect Count`, 
         -Groups, 
         -`Toxin producing species`,
         -`Microscope fields`) |> # remove columns that we won't use
  pivot_wider(names_from = Specie, # columns that contains the names of the new columns
              values_from = `Cells/L`, # column that contains the values for the new columns
              values_fn = max, 
              values_fill = 0) 
```


```{r echo=FALSE}
# assign each year to a custom color palette
# found the answer here: https://stackoverflow.com/questions/6919025/how-to-assign-colors-to-categorical-variables-in-ggplot2-that-have-stable-mappin
library(RColorBrewer)
years <- unique(z$year)
myColors <- brewer.pal(length(years),"Spectral")
names(myColors) <- years
```


# BIOTOXINS AND HABs PHYTOPLANKTON

## AST and *Pseudo-nitzschia sp.*

```{r message=FALSE, echo=FALSE}
# find the site-specific annual monthly mean for each of the pseudo-nitzschia groups counted
pn_monthly_mean <- z |>
  select(year, month, Station, `Pseudo-nitzschia sp.`, `Pseudo-nitzschia delicatissima gr.`, `Pseudo-nitzschia seriata gr.`) |>
  pivot_longer(cols = c(`Pseudo-nitzschia sp.`, `Pseudo-nitzschia delicatissima gr.`, `Pseudo-nitzschia seriata gr.`),
               names_to = "Species",
               values_to = "cells") |>
  group_by(Station, year, month) |>
  summarise("Cells/L" = mean(cells),
            sd = sd(cells, na.rm = TRUE),  # Standard deviation
            n = n(),                                        # Number of observations
            se = sd / sqrt(n),                              # Standard error
            .groups = "drop")
```


### Aquapark 1 (PA WB1) for oysters

```{r message=FALSE, echo=FALSE, fig.cap="Pseudo-nitzschia sp. Average cells/L/month in Aquapark 1 for 2018-2024", fig.width=8}
filter(pn_monthly_mean, Station == "Aquapark 1") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
  #ylim(0, 150000)
```

### Aquapark 4 (PA WB4) for oysters

```{r message=FALSE, echo=FALSE, fig.cap="Pseudo-nitzschia sp. Average cells/L/month in Aquapark 4 (pond 1) for 2018-2024", fig.width=8}
filter(pn_monthly_mean, Station == "Aquapark 4") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```


```{r message=FALSE, echo=FALSE, fig.cap="Pseudo-nitzschia sp. Average cells/L/month in Aquapark 4 (pond 2) for 2018-2024", fig.width=8}
filter(pn_monthly_mean, Station == "Nam Oysters SALT PANS") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```


### Lüderitz second lagoon (PA L2) for oysters

```{r echo=FALSE, message=FALSE}
# find the annual monthly mean for each of the pseudo-nitzschia groups counted at 3 combined Luderitz sites
lud_pn <- filter(z, Station %in% c("Luderitz Mariculture", 
                                   "NamOyster CC Lüderitz", 
                                   "Tetelestai Mariculture Lüderitz")) |>
  select(year, month, `Pseudo-nitzschia sp.`, `Pseudo-nitzschia delicatissima gr.`, `Pseudo-nitzschia seriata gr.`) |>
  pivot_longer(cols = c(`Pseudo-nitzschia sp.`, `Pseudo-nitzschia delicatissima gr.`, `Pseudo-nitzschia seriata gr.`),
               names_to = "Species",
               values_to = "cells") |>
  group_by(year, month) |>
  summarise("Cells/L" = mean(cells))
```

```{r message=FALSE, echo=FALSE, fig.cap="Pseudo-nitzschia sp. Average cells/L/month in Luderitz for 2018-2024", fig.width=8}
 ggplot(data = lud_pn, aes(x = month, y = `Cells/L`, fill = year))  +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```

### Lüderitz (PA L1) for abalone

```{r message=FALSE, echo=FALSE, fig.cap="Pseudo-nitzschia spp. cells-L (>0) – causative species for lipophillics at Luderitz abalone sites", fig.width=8}
filter(z, Station %in% c("Hangana Abalone")) |>
  select(`Date Collected`, year, month, `Pseudo-nitzschia sp.`, `Pseudo-nitzschia delicatissima gr.`, `Pseudo-nitzschia seriata gr.`) |>
  pivot_longer(cols = c(`Pseudo-nitzschia sp.`, `Pseudo-nitzschia delicatissima gr.`, `Pseudo-nitzschia seriata gr.`),
               names_to = "Species",
               values_to = "cells") |>
  filter(cells > 0) |>
  ggplot(aes(x = `Date Collected`, y = cells)) +
  geom_point() +
  theme_classic() +
  labs(x = "Date", y = "Cells/L")
```

## Lipophilic toxins (or DST) and *Dinophysis* spp

```{r message=FALSE, echo=FALSE}
# find the site-specific annual monthly mean for each of the dinophysis groups counted
dn_monthly_mean <- z |>
  select(year, month, Station, `Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`) |>
  pivot_longer(cols = c(`Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`),
               names_to = "Species",
               values_to = "cells") |>
  group_by(Station, year, month) |>
  summarise("Cells/L" = mean(cells))
```

### Aquapark 1 (PA WB1) for oysters

```{r message=FALSE, echo=FALSE, fig.cap="Dinophysis sp. Average cells/L/month in Aquapark 1 for 2018-2024", fig.width=8}
filter(dn_monthly_mean, Station == "Aquapark 1") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x=NULL, x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```


```{r warning=FALSE, fig.cap="Total Okadaic Acid Group test results in oysters for Aquapark 1. The Codex limit is 0.16 mg/kg."}
filter(y, Station %in% c("Fermar Seafoods CC (109)", "NamOysters cc Aquapark 1 (101)")) |>
  ggplot(aes(x = `Sampling Date`, y = okadaic_acid)) + 
  geom_point() + 
  geom_hline(yintercept = 0.16, color="red") +
  scale_y_continuous(breaks = seq(0,1,0.1), limits=c(0,1)) +
  theme_classic() +
  labs(x=NULL, y="Total Okadaic Acid Group mg/kg") 
```


```{r warning=FALSE, fig.cap="Total Yessotoxin (YTX) Group toxins in oysters from Aquapark 1."}
filter(y, Station %in% c("Fermar Seafoods CC (109)", "NamOysters cc Aquapark 1 (101)")) |>
  ggplot(aes(x = `Sampling Date`, y = `Yessotoxin value`)) + 
  geom_point() + 
  theme_classic() +
  labs(x=NULL, y="mg toxin /kg") 
```

### Aquapark 4 (PA WB4) for oysters

```{r message=FALSE, echo=FALSE, fig.cap="Dinophysis sp. Average cells/L/month in Aquapark 4 (pond 1) for 2018-2024", fig.width=8}
filter(dn_monthly_mean, Station == "Aquapark 4") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```

```{r message=FALSE, echo=FALSE, fig.cap="Mean Dinophysis spp cells/L in Aquapark 4 (pond 2) for 2018-2024", fig.width=8}
filter(dn_monthly_mean, Station == "Nam Oysters SALT PANS") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```


### Lüderitz second lagoon (PA L2) for oysters

```{r echo=FALSE, message=FALSE}
# find the annual monthly mean for each of the dinophysis groups counted at 3 combined Luderitz sites
lud_dn <- filter(z, Station %in% c("Luderitz Mariculture", 
                                   "NamOyster CC Lüderitz", 
                                   "Tetelestai Mariculture Lüderitz")) |>
  select(year, month, `Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`) |>
  pivot_longer(cols = c(`Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`),
               names_to = "Species",
               values_to = "cells") |>
  group_by(year, month) |>
  summarise("Cells/L" = mean(cells))
```


```{r message=FALSE, echo=FALSE, fig.cap="Dinophysis sp. Average cells/L/month in Luderitz for 2018-2024", fig.width=8}
 ggplot(data = lud_dn, aes(x = month, y = `Cells/L`, fill = year))  +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```



```{r message=FALSE, echo=FALSE, fig.cap="Dinophysis spp. cells-L – causative species for lipophillics at Luderitz oyster sites", fig.width=8}
filter(z, Station %in% c("Luderitz Mariculture", 
                         "NamOyster CC Lüderitz", 
                         "Tetelestai Mariculture Lüderitz")) |>
  select(`Date Collected`, year, month, `Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`) |>
  pivot_longer(cols = c(`Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`),
               names_to = "Species",
               values_to = "cells") |>
  filter(cells > 0) |>
  ggplot(aes(x = `Date Collected`, y = cells)) +
  geom_point() +
  theme_classic() +
  labs(x = "Date", y = "Cells/L")
```

###	Lüderitz (PA L1) for abalone 

```{r message=FALSE, echo=FALSE, fig.cap="Dinophysis spp. cells-L (>0) – causative species for lipophillics at Luderitz abalone sites", fig.width=8}
filter(z, Station %in% c("Hangana Abalone")) |>
  select(`Date Collected`, year, month, `Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`) |>
  pivot_longer(cols = c(`Dinophysis sp.`, `Dinophysis acuminata`, `Dinophysis fortii`, `Dinophysis accuta`),
               names_to = "Species",
               values_to = "cells") |>
  filter(cells > 0) |>
  ggplot(aes(x = `Date Collected`, y = cells)) +
  geom_point() +
  theme_classic() +
  labs(x = "Date", y = "Cells/L")
```


## PSTs and *Alexandrium* spp

```{r message=FALSE, echo=FALSE}
# find the site-specific annual monthly mean for each of the alexandrium groups counted
al_monthly_mean <- z |>
  select(year, month, Station, `Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, `Alexandrium catenella`, 
         `alexandrium Tamarense`) |>
  pivot_longer(cols = c(`Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, 
                        `Alexandrium catenella`, `alexandrium Tamarense`),
               names_to = "Species",
               values_to = "cells") |>
  group_by(Station, year, month) |>
  summarise("Cells/L" = mean(cells),
            sd = sd(cells, na.rm = TRUE),  # Standard deviation
            n = n(),                                        # Number of observations
            se = sd / sqrt(n),                              # Standard error
            .groups = "drop")
```


### Aquapark 1 (PA WB1) for oysters

```{r message=FALSE, echo=FALSE, fig.cap="Alexandrium sp. Average cells/L/month in Aquapark 1 for 2018-2024", fig.width=8}
filter(al_monthly_mean, Station == "Aquapark 1") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```


### Aquapark 4 (PA WB4) for oysters

```{r message=FALSE, echo=FALSE, fig.cap="Alexandrium sp. Average cells/L/month in Aquapark 4 for 2018-2024", fig.width=8}
filter(al_monthly_mean, Station == "Aquapark 4") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(name = "year", values = myColors)
```

```{r message=FALSE, echo=FALSE, fig.cap="Alexandrium sp. Average cells/L/month in Aquapark 4 (pond 2) for 2018-2024", fig.width=8}
filter(al_monthly_mean, Station == "Nam Oysters SALT PANS") |>
  ggplot(aes(x = month, y = `Cells/L`, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(name = "year", values = myColors)
```

### Lüderitz second lagoon (PA L2) for oysters

```{r echo=FALSE, message=FALSE}
# find the annual monthly mean for each of the alexandrium groups counted at 3 combined Luderitz sites
lud_al <- filter(z, Station %in% c("Luderitz Mariculture", 
                                   "NamOyster CC Lüderitz", 
                                   "Tetelestai Mariculture Lüderitz")) |>
  select(year, month, `Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, 
                        `Alexandrium catenella`, `alexandrium Tamarense`) |>
  pivot_longer(cols = c(`Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, 
                        `Alexandrium catenella`, `alexandrium Tamarense`),
               names_to = "Species",
               values_to = "cells") |>
  group_by(year, month) |>
  summarise("Cells/L" = mean(cells))
```


```{r message=FALSE, echo=FALSE, fig.cap="Alexandrium sp. Average cells/L/month in Luderitz for 2018-2024", fig.width=8}
 ggplot(data = lud_al, aes(x = month, y = `Cells/L`, fill = year))  +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_discrete(labels = function(x) month.abb[as.numeric(x)]) +
  labs(x = "Month", y = "Cells/L", fill = "Year") +
  theme_classic() +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.title = element_blank()) +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(name = "year", values = myColors)
```


```{r message=FALSE, echo=FALSE, fig.cap="Alexandrium spp. cells-L (>0) – causative species for PSTs at Luderitz oyster sites", fig.width=8}
filter(z, Station %in% c("Luderitz Mariculture", 
                         "NamOyster CC Lüderitz", 
                         "Tetelestai Mariculture Lüderitz")) |>
  select(`Date Collected`, year, month, `Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, 
                        `Alexandrium catenella`, `alexandrium Tamarense`) |>
  pivot_longer(cols = c(`Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, 
                        `Alexandrium catenella`, `alexandrium Tamarense`),
               names_to = "Species",
               values_to = "cells") |>
  filter(cells > 0) |>
  ggplot(aes(x = `Date Collected`, y = cells)) +
  geom_point() +
  theme_classic() +
  labs(x = "Date", y = "Cells/L")
```


```{r message=FALSE, echo=FALSE, fig.cap="Alexandrium spp. cells-L (>0) – causative species for PSTs at Luderitz abalone sites", fig.width=8}
filter(z, Station %in% c("Hangana Abalone")) |>
  select(`Date Collected`, year, month, `Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, 
                        `Alexandrium catenella`, `alexandrium Tamarense`) |>
  pivot_longer(cols = c(`Alexandrium sp.`, `Alexandrium minutum`, `Alexandrium Tamarense`, 
                        `Alexandrium catenella`, `alexandrium Tamarense`),
               names_to = "Species",
               values_to = "cells") |>
  filter(cells > 0) |>
  ggplot(aes(x = `Date Collected`, y = cells)) +
  geom_point() +
  theme_classic() +
  labs(x = "Date", y = "Cells/L")
```