---
title: "THE NAMIBIAN SHELLFISH SANITATION PROGRAMME (NSSP) REVIEW REPORT FOR 2024/25"
output: word_document
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r}
source("../setup.R")
```

```{r}
source("../code/19_mbdata_functions.R")
```


```{r}
mbdata <- read_microbio_data()
```

# MICROBIOLOGICAL PARAMETERS

The biological agents involved in seafood contamination include bacteria, viruses, and parasites, which can cause illnesses ranging from mild gastroenteritis to life-threatening diseases. Some of these pathogens are naturally present in the aquatic environment, while others are introduced through animal or human fecal shedding and sewage pollution. Bacteria naturally present in seawater are generally found in limited numbers in live and raw seafood; however, filter-feeding molluscan shellfish can concentrate these microorganisms, posing a significant risk since they are often consumed raw.

## *Escherichia coli*

The Category I limit for E. coli is 230 MPN/ 100 grams of flesh. The detection limit of the test method is 18 MPN/ 100 grams of flesh.

### Aquapark 1 (PA WB1) for oysters

```{r fig.cap="*E. coli* results in oysters at Aquapark 1(PA WB1). Red line represents Category I limit"}
plot_ecoli(mbdata, station = c("Fermar Seafoods CC (109)"))
```


### Aquapark 4 (PA WB4) for oysters

Farm 112 (Pond 1)

```{r fig.cap="*E. coli* results in oysters at Aquapark 4 (112). Red line represents the Category I limit."}
plot_ecoli(mbdata, station = c("Aquapark 4 (112)"))
```

Farm 115 (Pond 2)

```{r fig.cap="*E. coli* results in oysters at Aquapark 4 (115). Red line represents the Category I limit."}
plot_ecoli(mbdata, station = c("Nam Oysters SALT PANS (115)", "Nam Oysters Salt Pans (115)"))
```


### Lüderitz lagoon (PA L2) for oysters

```{r fig.cap="*E. coli* results in oysters at Lüderitz (PA L2). Red line represents Category I limit."}
plot_ecoli(mbdata, station = c("Luderitz Mariculture (Pty) Ltd (105)",
                               "Five Roses cc (106)",
                               "Lagoon aquaculture CC (111)",
                               "NamOyster CC Lüderitz (121)", 
                               "Tetelestai Mariculture Lüderitz (120)",
                               "Tetelestai Mariculture cc"))
```


### Lüderitz (PA L1) for abalone

```{r fig.cap="*E. coli* results in for abalone at Lüderitz (PA L1). Red line represents Category I limit."}
plot_ecoli(mbdata, station = c("Hangana Abalone (103)"))
```


## *Vibrio* spp.

*Vibrio* spp. results are reported qualitatively as either "present" or "absent" per 25 g of flesh. Until August 2024, *Vibrio parahaemolyticus* was reported only qualitatively (as present or absent). From September 2024 onward, *V. parahaemolyticus* has been reported both qualitatively and quantitatively.

### Aquapark 1 (PA WB1) for oysters

```{r}
plot_vibrio(mbdata, station = c("Fermar Seafoods CC (109)"))
```

```{r}
plot_vibrio_par(mbdata, station = c("Fermar Seafoods CC (109)"))
```


### Aquapark 4 (PA WB4) for oysters

Farm 112 (Pond 1)

```{r}
plot_vibrio(mbdata, station = c("Aquapark 4 (112)"))
```

```{r}
plot_vibrio_par(mbdata, station = c("Aquapark 4 (112)"))
```

Farm 115 (Pond 2)

```{r}
plot_vibrio(mbdata, station = c("Nam Oysters SALT PANS (115)", "Nam Oysters Salt Pans (115)"))
```

```{r}
plot_vibrio_par(mbdata, station = c("Nam Oysters SALT PANS (115)", "Nam Oysters Salt Pans (115)"))
```


### Lüderitz lagoon (PA L2) for oysters

```{r}
plot_vibrio(mbdata, station = c("Luderitz Mariculture (Pty) Ltd (105)",
                                "Five Roses cc (106)",
                                "Lagoon aquaculture CC (111)",
                                "NamOyster CC Lüderitz (121)", 
                                "Tetelestai Mariculture Lüderitz (120)",
                                "Tetelestai Mariculture cc"))
```

```{r}
plot_vibrio_par(mbdata, station = c("Luderitz Mariculture (Pty) Ltd (105)",
                                    "Five Roses cc (106)",
                                    "Lagoon aquaculture CC (111)",
                                    "NamOyster CC Lüderitz (121)", 
                                    "Tetelestai Mariculture Lüderitz (120)",
                                    "Tetelestai Mariculture cc"))
```

### Lüderitz (PA L1) for abalone

```{r}
plot_vibrio(mbdata, station = c("Hangana Abalone (103)"))
```

```{r}
plot_vibrio_par(mbdata, station = c("Hangana Abalone (103)"))
```


## *Salmonella*


