---
title: "Data Analysis and Visualisation"
author: "Johnathan Evanilla"
date: "10/28/2024"
output: github_document
---

First, we'll need to load some packages. 

```{r}
#| include=FALSE
library(readxl)
library(dplyr)
library(ggplot2)
```

R comes with lots of great tools out of the box, but there are others that people develop for specific tasks.



# Reading files with data

Set your path to the "Qry_Phytoplankton Total Cell_L DATA.XLSX" spreadsheet in quotes below:

```{r}
my_path <- file.path("~/Documents/Bigelow/data/Qry_Phytoplankton Total Cell_L DATA.XLSX")
```


The package `readxl` provides tools for reading excel files (ie .xls and .xlsx) 

```{r}
x <- read_excel(path = my_path)

x
```


The `readr` package also has functions for reading other file types like:

`read_csv()` - comma separated
`read_tsv()` - tab separated

These files can be stored locally on you computer or hosted at a url like https://



## Inspecting data.frame objects

We already saw how the functions head() and str() can be useful to check the content and the structure of a data.frame. Here is a non-exhaustive list of functions to get a sense of the content/structure of the data.

### Size:
- `dim()` - returns a vector with the number of rows in the first element, and the number of columns as the second element (the dimensions of the object)
- `nrow()` - returns the number of rows
- `ncol()` - returns the number of columns

```{r}
dim(x)
nrow(x)
ncol(x)
```

### Content:
- `head()` - shows the first 6 rows
- `tail()` - shows the last 6 rows

```{r}
head(x, 10)
tail(x)
```

### Names:
- `names()` - returns the column names (synonym of colnames() for data.frame objects)

```{r}
names(x)
```

### Summary:

- `str()` - structure of the object and information about the class, length and content of each column
- `summary()` - summary statistics for every column in the data.frame
- `glimpse()` - similar to `str()` but shows you more rows of data across the screen; the table is transposed sideways

```{r}
str(x)
```

```{r}
summary(x)
```

`summary()` can also be called on a single column

```{r}
summary(x$`Cells/L`)
```

```{r}
glimpse(x)
```

Note: most of these functions are “generic”, meaning they can be used on other types of objects beyond a data.frame.

`View()` also allows you to open the data.frame in a new window to explore by hand

## Working on data frames using dplyr

We can print the data.frame to look at by just calling its name

Notice that all of the columns don't fit onto the page, but you can use the right arrow to see the others

```{r}
x
```


### Subsetting columns

`select()` lets us subset specified columns

```{r}
select(x, `Date Collected`, Station, Specie, `Cells/L`)
```

The pipe operator `|>` passes the result of the content on its left as the first argument to the function on its right

The code below yields the same result as the code just above

```{r}
x |>
  select(`Date Collected`, Station, Specie, `Cells/L`)
```

### Filtering values

`filter()` is for getting a subset of rows based on a condition


```{r}
x |>
  filter(`Cells/L` < 100)
```

The pipe operator is useful for chaining together operations on our data.frame

```{r}
x |> 
  select(`Date Collected`, Station, Specie, `Cells/L`) |>
  filter(`Cells/L` < 100)
```

Subsetting by character string

```{r}
x |>
  filter(Station == "Production Area 1")
```

### Mutating new columns

`mutate()` - change or make new columns 

```{r}
x |>
  mutate(Groups = tolower(Groups))
```


```{r}
x |>
  mutate(id = paste(`Date Collected`, Station, Specie, sep="_"))
```

Let's add a month and week column

```{r}
x <- x |>
  mutate(month = format(`Date Collected`, format="%m"),
         week = format(`Date Collected`, format="%U"),
         year = format(`Date Collected`, format="%Y"))
```

`format()` is useful for modifying date formats

There are lot's of format codes for dates in R. [list](https://www.r-bloggers.com/2013/08/date-formats-in-r/) contains a lot of them.

```{r}
format(x$`Date Collected`[1], format="%b")
```


### Renaming columns

`rename()` will change the name of columns

Use the syntax `new_name = old_name` in the arguments

```{r}
rename(x, date = `Date Collected`)
```

### Ordering rows

`arrange()` updates the row order by a column

```{r}
arrange(x, `Cells/L`)
```

`desc()` will order rows based on descending order of the column

```{r}
arrange(x, desc(`Cells/L`))
```


# *Alexandrium* data


```{r}
al <- filter(x, Specie %in% c("Alexandrium sp.", "Alexandrium Tamarense", "Alexandrium acatenella", "Alexandrium catenella", "Alexandrium minutum", "alexandrium Tamarense"))
```


```{r}
al
```

`count()` let's us see the unique values in a column of the data.frame

```{r}
count(al, Station)
```


## Base R plotting

We can make histograms using `hist()`

```{r}
hist(al$`Cells/L`)
```

The breaks argument lets us specify number of breaks in the plot

```{r}
hist(al$`Cells/L`, breaks=25)
```

Log-scaled histogram

```{r}
hist(log10(al$`Cells/L`+1))
```

The `plot()` function can be used for simple plotting of two variables

```{r}
plot(`Cells/L`~month, data=al)
```


## Plotting with `ggplot2`

We can make plots by creating a plot object with `ggplot()`, then use the `+` operator to add features to it

Here's the same plot as above but made with `ggplot2`

```{r}
ggplot(data=al, aes(x=month, y=`Cells/L`)) +
  geom_point(alpha=0.3)
```

We can make a subset table of the data to use in plots that focus on more specific data

```{r}
y <- al |>
  filter(year == 2019,
         Station == "Production Area 2")

y
```

```{r}
ggplot(data = al, aes(x=`Date Collected`, y=`Cells/L`)) +
  geom_line()
```


Here's a line plot of *Alexandrium* abundance at one site during one year

```{r}
ggplot(y, aes(x=`Date Collected`, y = `Cells/L`)) +
  geom_line()
```



```{r}
ggplot(data=al, aes(x=week, y=Station, fill=`Cells/L`)) +
  geom_tile() +
  scale_fill_viridis_b() +
  theme_bw()
```


We can use `facet_wrap()` or `facet_grid()` to make multiple plots based on a grouping variable in our data


```{r}
al |>
  filter(Station %in% c("Production Area 1", "Production Area 2", "Production Area 3")) |>
  ggplot(aes(x=`Date Collected`, y=`Cells/L`)) +
  geom_line() +
  facet_wrap(vars(Station))
```

The code above produces a warning because every station doesn't have data

```{r}
al |>
  filter(Station %in% c("Production Area 1", "Production Area 2", "Production Area 3")) |>
  ggplot(aes(x=`Date Collected`, y=`Cells/L`)) +
  geom_line() +
  facet_wrap(vars(Station))
```

# *Dinophysis* data

```{r}
dn <- filter(x, Specie == "Dinophysis acuminata")
```

```{r}
dn <- filter(x, Specie %in% c("Dinophysis acuminata", "Dinophysis fortii", "Dinophysis sp.", "Dinophysis rotundata", "Dinophysis accuta"))
```


```{r}
count(dn, Station)
```

```{r}
hist(dn$`Cells/L`)
```

```{r}
hist(log10(dn$`Cells/L`))
```

```{r}
plot(`Cells/L`~month, data=dn)
```

```{r}
ggplot(data=dn, aes(x=week, y=Station, fill=`Cells/L`)) +
  geom_tile() +
  scale_fill_viridis_b() +
  theme_bw()
```

# *Pseudo-nitzschia* data

```{r}
pn <- filter(x, Specie %in% c("Pseudo-nitzschia sp.", "Pseudo-nitzschia spp.", "Pseudo-nitzschia seriata gr.", "Pseudo-nitzschia delicatissima gr."))
```


```{r}
plot(`Cells/L`~month, data=pn)
```

```{r}
ggplot(data=pn, aes(x=`Date Collected`, y=Station, fill=`Cells/L`)) +
  geom_tile() +
  scale_fill_viridis_b() +
  theme_bw()
```


```{r}
x |>
  filter(`Toxin producing species` %in% c("PSP", "ASP", "DSP")) |>
  ggplot(aes(x=week, y=Station, fill=`Toxin producing species`)) +
  geom_tile()
```

