---
title: "Introduction to Programming"
author: "Johnathan Evanilla"
date: "10/28/2024"
output: github_document
---

## Welcome

## Rstudio Tour

- Console Pane

- Environment Pane

- Text Editor Pane

- File Browser Pane


### Hello World!

```{r}
print("Hello world!")
```


```{r}
getwd()
```


## Creating Objects

```{r}
3+3
```

```{r}
# I'm adding 3 and 5
3+5
```


```{r error=TRUE}
#I'm adding 3 and 5
3+5
```

```{r}
# assign 3 to a
a <- 3
# assign 5 to b
b <- 5

# what now is a
a
# what now is b
b

#Add a and b
a + b
```


## Functions

# Create a function that adds two numbers

```{r}
adder <- function(a,b) {
  c = a+b
  return(c)
}
```

# test if it works

```{r}
adder(5,5)
```

# R has lots of functions built in, like `round()`

```{r}
round(3.14159)
```

# typing a `?` in front of the function will access the help page for that function

```{r}
?round
```

# try using the `digits`

```{r}
round(3.14159, digits=2)
```

# you don't need to declare the argument name, R can figure that out for you

```{r}
round(3.14159, 2)
```


## Vectors and data types

A vector of doubles

```{r}
counts <- c(1, 2, 3, 100)
counts
```

A vector of character

```{r}
species <- c("alexandrium", "dinophysis", "pseudo_nitzchia")
species
```


```{r}
class(counts)
class(species)
```

```{r}
numbers <- c("1", "2", "3")
class(numbers)
```

R also has a "date" class

```{r}
today <- "2022-12-07"
```



```{r}
class(today)
```


We can use the function `as.Date()` to change to date class
```{r}
today <- as.Date(today)
class(today)
```

[Date Formats in R](https://www.r-bloggers.com/2013/08/date-formats-in-r/)
```{r}
format(today, format="%A") #day of week
format(today, format="%B") #month
format(today, format="%U") #week of year
```


```{r}
Sys.Date()
class(Sys.Date())
```


The logical class is used to store binary variables i.e. true or false,

```{r}
present <- TRUE
class(present)
```


```{r}
length(counts)
length(species)
```

```{r}
2 * counts
```

Notice that you can't add character and numeric vectors together

```{r}
# counts + species
```


```{r}
class(counts)
class(species)
```

```{r}
str(counts)
str(species)
```

```{r}
counts <- c(counts, 4) # adding at the end
counts <- c(0, counts) # adding at the beginning
counts
```


```{r}
counts[3]
```



```{r}
counts[-1]
```

```{r}
mean(counts)
```


```{r}
sd(counts)
```

```{r}
sessionInfo()
```


Content adapted from https://datacarpentry.org/R-genomics/01-intro-to-R.html


